Follow-up Samples

### 2017-07-22 submission/manuscript.tex
add back old version

\usepackage{lineno}

\usepackage{pdfpages}
\usepackage{comment}
\usepackage{lscape}

@@

\textbf{\emph{Difficult to identify effects of specific treatments on
the change in the microbiota.}} The type of treatment that the patients
received varied across diagnosis groups. Those with adenomas and
advanced adenomas received surgical resection (adenoma, N=4; advanced
adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=18;
advanced adenoma, N=15) and those with carcinomas received surgical
resection (N=12), surgical resection with chemotherapy (N=9), and
surgical resection with chemotherapy and radiation (N=5). We focused on
the patients with carcinoma and pooled those patients that received
chemotherapy with those that received chemotherapy and radiation to
improve our statistical power. We did not observe a significant
difference in the effect of these treatments on the number of observed
OTUs, Shannon diversity, or Shannon evenness (P-value \textgreater{}
0.05). Furthermore, there was not a significant difference in the effect
of the treatments on the amount of change in the community structure
(P-value = 0.298). Finally, the change in the positive probability was
not significantly different between the two treatment groups (P-value =
0.999). Due to the relatively small number of samples in each treatment
group, it was difficult to make a definitive statement regarding the
specific type of treatment on the amount of change in the structure of
the microbiota.

@@

Our study focused on comparing the microbiota of patients diagnosed with
adenoma, advanced adenoma, and carcinoma before and after treatment. For
all three groups of patients, we observed changes in their microbiota.
After treatment, the microbiota of patients with carcinoma changed
significantly more than the other groups. This change resulted in
communities that more closely resembled those of patients with a normal
colon. This may suggest that treatment for carcinoma is not only
successful for removing the carcinoma but also at reducing the
associated bacterial communities. Understanding the effect of treatment
on the microbiota of those diagnosed with carcinomas may have important
implications for reducing disease recurrence. It is intriguing that it
may be possible to use microbiome-based biomarkers to not only predict
the presence of lesions but to assess the risk of recurrence.

@@

This study expands upon existing research that has established a role
for the microbiota in tumorigenesis and that demonstrated the utility of
microbiome-based biomarkers to predict the presence of colonic lesions.
The most exciting future direction from the current study is the
possibility that markers within the microbiota could be used to evaluate
the effect of treatment and predict recurrence for those diagnosed with
carcinoma. If such an approach is effective, it might be possible to
target the microbiota as part of adjuvant therapy. Our data provides
additional evidence on the importance of the microbiota in tumorigenesis
by addressing the recovery of the microbiota after treatment and opens
interesting avenues of research into how these changes may affect
recurrence.

@@

\textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
have been previously reported in Baxter, et al {[}24{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between one and two weeks after initial colonoscopy. The bacterial
community has been shown to normalize back to a pre-colonoscopy
community within this time period {[}27{]}. Our study cohort consisted
of 67 individuals with an initial sample as described and a follow up
sample obtained between 188 - 546 days after treatment of lesion
{[}Table 1{]}. Patients were diagnosed by colonoscopic examination and
histopathological review of any biopsies taken. Patients were classified
as having advanced adenoma if they had an adenoma greater than 1 cm,
more than three adenomas of any size, or an adenoma with villous
histology. This study was approved by the University of Michigan
Institutional Review Board. All study participants provided informed
consent and the study itself conformed to the guidelines set out by the
Helsinki Declaration.

@@

Similarily, the Random Forest {[}33{]} algorithm was also used to create
the three models used to classify normal versus diagnosis. The total
number of individuals in the normal versus diagnosis models was 423
individuals {[}Table 2{]}. There were a total of 239 individuals in the
normal versus adenoma model, 262 individuals in the normal versus
advanced adenoma model, and 266 individuals in the normal versus
carcinoma model {[}Table 2{]}.

All models included only OTU data obtained from 16S rRNA sequencing and
were processed and cleaned using the R package caret (v6.0.73).
Optimization of the mtry hyper-parameter involved making 100 different
80/20 (train/test) splits of the data where the same proportion was
present within both the whole data set and the 80/20 split. For each of
the different splits, 20 repeated 10-fold cross validation was performed
on the 80\% component to optimize the mtry hyper-parameter by maximizing
the AUC (Area Under the Curve of the Receiver Operator Characteristic).
The resulting model was then tested on the hold out data obtained from
the 20\% component. For all pre versus post-treatment models the
optimized mtry was 2 and for all normal versus diagnosis models the
optimized mtry was 2. The hyper-parameter, mtry, defines the number of
variables to investigate at each split before a new division of the data
was created with the Random Forest model {[}33{]}.

@@

\textbf{\emph{Statistical Analysis.}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}34{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall
differences in the median relative abundance of each OTU between pre and
post-treatment samples utilized a paired Wilcoxon ranked sum test. Where
multiple comparison testing was appropriate, a Benjamini-Hochberg (BH)
correction was applied {[}35{]} and a corrected P-value of less than
0.05 was considered significant. The P-values reported are those that
were BH corrected. Model rank importance was determined by obtaining the
median MDA from the 100, 20 repeated 10-fold cross validation and then
ranking from largest to smallest MDA.

@@

\textbf{Figure 2: The 10 OTUs used to classify treatment for adenoma,
advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
OTUs. C) Carcinoma OTUs.

@@

\begin{longtable}[]{@{}lccc@{}}
\toprule
& Adenoma & Advanced Adenoma & Carcinoma\tabularnewline
\midrule
\endhead
n & 22 & 19 & 26\tabularnewline
Age (Mean ± SD) & 61.68 ± 7.2 & 63.11 ± 10.9 & 61.65 ±
12.9\tabularnewline
Sex (\%F) & 36.36 & 36.84 & 42.31\tabularnewline
BMI (Mean ± SD) & 26.86 ± 3.9 & 25.81 ± 4.7 & 28.63 ± 7.2\tabularnewline
Caucasian (\%) & 95.45 & 84.21 & 96.15\tabularnewline
Days Between Colonoscopy (Mean ± SD) & 255.41 ± 42 & 250.16 ± 41 &
350.85 ± 102\tabularnewline
\bottomrule
\end{longtable}

>>>>

\usepackage{lineno}
\usepackage{microtype}
\usepackage{pdfpages}
\usepackage{comment}
\usepackage{lscape}

@@

\textbf{\emph{Difficult to identify effects of specific treatments on
the change in the microbiota.}} The type of treatment that the patients
received varied across diagnosis groups. Those with adenomas and
advanced adenomas received surgical resection (adenoma, N=4; advanced
adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=18;
advanced adenoma, N=15) and those with carcinomas received surgical
resection (N=12), surgical resection with chemotherapy (N=9), and
surgical resection with chemotherapy and radiation (N=5). We focused on
the patients with carcinoma and pooled those patients that received
chemotherapy with those that received chemotherapy and radiation to
improve our statistical power. We did not observe a significant
difference in the effect of these treatments on the number of observed
OTUs, Shannon diversity, or Shannon evenness (P-value \textgreater{}
0.05). Furthermore, there was not a significant difference in the effect
of the treatments on the amount of change in the community structure
(P-value = 0.532). Finally, the change in the positive probability was
not significantly different between the two treatment groups (P-value =
0.532). Due to the relatively small number of samples in each treatment
group, it was difficult to make a definitive statement regarding the
specific type of treatment on the amount of change in the structure of
the microbiota.

@@

Our study focused on comparing the microbiota of patients diagnosed with
adenoma, advanced adenoma, and carcinoma before and after treatment. For
all three groups of patients, we observed changes in their microbiota.
Some of these changes, specifically for adenoma and advanced adenoma,
may be due to normal temporal variation but those with CRC clearly had
large microbiota changes that would be greater than what could be
expected from this type of variation. After treatment, the microbiota of
patients with carcinoma changed significantly more than the other
groups. This change resulted in communities that more closely resembled
those of patients with a normal colon. This may suggest that treatment
for carcinoma is not only successful for removing the carcinoma but also
at reducing the associated bacterial communities. Understanding the
effect of treatment on the microbiota of those diagnosed with carcinomas
may have important implications for reducing disease recurrence. It is
intriguing that it may be possible to use microbiome-based biomarkers to
not only predict the presence of lesions but to assess the risk of
recurrence.

@@

This study expands upon existing research that has established a role
for the microbiota in tumorigenesis and that demonstrated the utility of
microbiome-based biomarkers to predict the presence of colonic lesions.
Although we cannot make a specific statment regarding the effect of a
specific treatment on the gut microbiota, due to the low number of
samples, there are still a number of exciting new avenues to explore.
One of the most exciting of these future directions is the possibility
that markers within the microbiota could be used to potentially evaluate
the effect of treatment and predict recurrence for those diagnosed with
carcinoma. If such an approach is effective, it might be possible to
target the microbiota as part of adjuvant therapy, if the biomarkers
identified play a key role in the disease process. Our data provides
additional evidence on the importance of the microbiota in tumorigenesis
by addressing the recovery of the microbiota after treatment and opens
interesting avenues of research into how these changes may affect
recurrence.

@@

\textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
have been previously reported in Baxter, et al {[}24{]}. Briefly,
samples were stored on ice for at least 24h before freezing. Although we
cannot exclude that this sampling protocol may have impacted the gut
microbiota composition all samples were subjected to the same
methodology. Study exclusion involved those who had already undergone
surgery, radiation, or chemotherapy, had colorectal cancer before a
baseline fecal sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or familial adenomatous polyposis.
Samples used to build the models for prediction were collected either
prior to a colonoscopy or between one and two weeks after initial
colonoscopy. The bacterial community has been shown to normalize back to
a pre-colonoscopy community within this time period {[}27{]}. Our study
cohort consisted of 67 individuals with an initial sample as described
and a follow up sample obtained between 188 - 546 days after treatment
of lesion {[}Table 1{]}. Patients were diagnosed by colonoscopic
examination and histopathological review of any biopsies taken. Patients
were classified as having advanced adenoma if they had an adenoma
greater than 1 cm, more than three adenomas of any size, or an adenoma
with villous histology. This study was approved by the University of
Michigan Institutional Review Board. All study participants provided
informed consent and the study itself conformed to the guidelines set
out by the Helsinki Declaration.

\textbf{\emph{Treatment}} The majority of patients undergoing treatment
for adenoma or advanced adenoma were not treated surigically {[}Table
1{]}. The type of chemotherapy used for patients with CRC varied
markably. The types used included Oxaliplatin, Levicovorin, Folfox,
Xeloda, Capecitabine, Avastin, Fluorouracil, and Glucovorin. These were
used individually or in combination with others depending on the patient
{[}Table 1{]}. If an individual was treated with radiation they were
also always treated with chemotherapy. Radiation therapy generally used
18 mV photons for treatment.

@@

Similarily, the Random Forest {[}33{]} algorithm was also used to create
the three models used to classify normal versus diagnosis. These samples
were obtained using the same methodology as described earlier in this
section. All samples used for this component of model training were from
pre-treatment samples. The total number of individuals in the normal
versus diagnosis models was 423 individuals {[}Table 2{]}. There were a
total of 239 individuals in the normal versus adenoma model, 262
individuals in the normal versus advanced adenoma model, and 266
individuals in the normal versus carcinoma model {[}Table 2{]}.

All models included only OTU data obtained from 16S rRNA sequencing and
were processed and cleaned using the R package caret (v6.0.76).
Optimization of the mtry hyper-parameter involved making 100 different
80/20 (train/test) splits of the data where the same proportion was
present within both the whole data set and the 80/20 split. For each of
the different splits, 20 repeated 10-fold cross validation was performed
on the 80\% component to optimize the mtry hyper-parameter by maximizing
the AUC (Area Under the Curve of the Receiver Operator Characteristic).
The resulting model was then tested on the hold out data obtained from
the 20\% component. For all pre versus post-treatment models the
optimized mtry was 2 and for all normal versus diagnosis models the
optimized mtry was 2. The hyper-parameter, mtry, defines the number of
variables to investigate at each split before a new division of the data
was created with the Random Forest model {[}33{]}.

@@

\textbf{\emph{Statistical Analysis.}} The R software package (v3.4.1)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}34{]} in the vegan package
(v2.4.3). Comparisons between probabilities as well as overall
differences in the median relative abundance of each OTU between pre and
post-treatment samples utilized a paired Wilcoxon ranked sum test. Where
multiple comparison testing was appropriate, a Benjamini-Hochberg (BH)
correction was applied {[}35{]} and a corrected P-value of less than
0.05 was considered significant. The P-values reported are those that
were BH corrected. Model rank importance was determined by obtaining the
median MDA from the 100, 20 repeated 10-fold cross validation and then
ranking from largest to smallest MDA.

@@

\textbf{Figure 2: The 10 OTUs used to classify treatment for adenoma,
advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
OTUs. C) Carcinoma OTUs. The darker circle highlights the median log10
MDA value obtained from 100 different 80/20 splits while the lighter
colored circles represents the value obtained for a specific run.

@@

\begin{longtable}[]{@{}lccc@{}}
\toprule
& Adenoma & Advanced Adenoma & Carcinoma\tabularnewline
\midrule
\endhead
n & 22 & 19 & 26\tabularnewline
Age (Mean ± SD) & 61.68 ± 7.2 & 63.11 ± 10.9 & 61.65 ±
12.9\tabularnewline
Sex (\%F) & 36.36 & 36.84 & 42.31\tabularnewline
BMI (Mean ± SD) & 26.86 ± 3.9 & 25.81 ± 4.7 & 28.63 ± 7.2\tabularnewline
Caucasian (\%) & 95.45 & 84.21 & 96.15\tabularnewline
Days Between Colonoscopy (Mean ± SD) & 255.41 ± 42 & 250.16 ± 41 &
350.85 ± 102\tabularnewline
Surgery Only & 4 & 4 & 12\tabularnewline
Surgery \& Chemotherapy & 0 & 0 & 9\tabularnewline
Surgery, Chemotherapy, \& Radiation & 0 & 0 & 5\tabularnewline
\bottomrule
\end{longtable}
